---
title: "Hypothesis Testing: Exercises"
author: ""
date: "September 21/24, 2020"
output: 
  html_document:
    self_contained: yes
---

This document is meant to be used to practice after you have completed the tutorial for today's lab. You can find a completed version of these exercises at <https://jdbest.github.io/psychRstats/answers.html>

If you would like to refer to the tutorial while also working on this document, there are instructions on the [wiki](https://github.com/jdbest/psychRstats/wiki/Run-a-tutorial-and-exercise-simultaneously) on how to do so.

You may also want to refer to past labs. Don't forget that previous labs are linked to on the main [labs website](https://jdbest.github.io/psychRstats/labs.html).

## Objectives

In the tutorial, you put some of the work we did in class into practice in exploring a hypothesis test. Instead of using the `friends` data we focused on in the tutorial, we'll be using some real data for this example. 

I will again encourage you to do two things as you work through this document: (1) Save it frequently! Hit the disk image above to save, hit Command/Ctrl and s, or go to the File menu and click save. When the document is saved, the title of it will go from red to black. (2) Practice Knitting the document by hitting the Knit button. You can do it now---it'll create an HTML file in the folder where this file lives. Knitting will help to identify problems and provide a complete document for sharing.

## Loading packages

Run the following code chunk to load necessary packages for these exercises. 

```{r loadpackages}
library(tidyverse)
# tidyverse loads these:
# library(dplyr)
# library(ggplot2)
```

## Practicing a hypothesis test

We'll return to importing new data next week---for today, just run the following chunk to load the data.

```{r}
tvfriends <- structure(list(tvhours = c(4, 1, 0, 0, 5, 4, 0, 0, 15, 8, 2, 
4, 0, 5, 4, 3, 7, 5, 14, 0, 6, 3, 0, 8, 0, 1, 10, 14, 15, 6, 
4.5, 17, 7, 10, 7, 3, 2, 14, 5, 5, 2, 2, 2, 7, 7, 4, 7, 10, 3, 
5, 20, 14, 1, 8, 3, 7, 0.5, 8, 3, 1, 20, 2, 1, 15, 14, 5, 1, 
0, 4), gender = c("female", "female", "male", "female", "female", 
"female", "female", "non-binary", "female", "male", "female", 
"female", "agender", "male", "male", "female", "female", "female", 
"female", "female", "female", "male", "male", "female", "male", 
"male", "female", "male", "female", "male", "male", "female", 
"female", "female", "male", "male", "female", "female", "non-binary", 
"male", "female", "male", "female", "male", "female", "female", 
"male", "male", "female", "male", "female", "female", "female", 
"female", "female", "female", "female", "female", "female", "female", 
"male", "female", "male", "female", "male", "female", "female", 
"male", "male")), row.names = c(NA, -69L), class = c("tbl_df", 
"tbl", "data.frame"))
```

As you did in the tutorial, you're going to walk through the five steps of hypothesis-testing, calculate the *z*-score, determine the cut-off value, and get a *p*-value. 

## Would someone who doesn't watch any TV be considered a statistically-significant outlier among Bard students? 

Take a look at the structure of your data using the str() function or by clicking on the data in the Environment pane. 

1. Restate question as a research and null hypothesis

Write the research and null hypotheses, and then frame them in terms of means.

> (Yes, delete this text and write your answer after the greater than sign.)

Before we continue: what's **your** hypothesis? Do you think 0 hours fits with most Bard students? 

> 

2. Determine the characteristics of the comparison distribution

Describe the z-distribution.

> 

3. Determine the sample cutoff score to reject the null hypothesis. 

First, what cutoff do you expect for a two-tailed distribution of z-scores?

> 

Then, find the code from the tutorial. Run it here.

```{r ex3-cutoff}

```

4. Determine your sample’s score

Plot a boxplot of the data, and a histogram.

```{r ex4}

```

Remember, the question was "How many hours of Netflix/Hulu/TV do you watch per week?" Do any of the data seem unreasonable? Should you filter it? Describe the histogram. 

> 

```{r filter}

```

Calculate the mean and standard deviation, and save them. Calculate your `z` for someone who watches 0 hours of TV a week. (`x = 0`)

```{r ex4-calc}

```

5. Decide whether or not to reject the null hypothesis

Based on the cut-off defined in Step 3, should we reject the null? 

> 

What conclusion should you draw based on that?

> 

Try just running the `pnorm()` function on the *z*-value. Then, below the code chunk, explain what percentile this person is in.

```{r ex5}

```

> 

## Part II: Work through the steps on your own for a one-sided test

Would that score (0 `tvhours`) be statistically significant if we're only interested in students who watch significantly **less** TV than other Bard students? Just write the responses below this line, and use code as necessary. 



Once you've completed the lab, knit it.

## Extension exercise

If you have more time, want to practice, or are just interested, this is another exercise of the same vein. 

This is a summary of data from *Pro Publica*, an investigative journalism organization. You can read more about the data here: <https://www.propublica.org/article/so-sue-them-what-weve-learned-about-the-debt-collection-lawsuit-machine>

|Type              |     Mean|      SD|
|:-----------------|--------:|-------:|
|Auto              |   109.52|  236.87|
|Collection Agency |    39.26|   93.39|
|Debt Buyer        |   316.98|  867.30|
|Government        |   121.04|  190.46|
|High-Cost Lender  |    46.00|   53.49|
|Insurance         |   128.13|  218.79|
|Major Bank        |   690.02| 1440.31|
|Medical           |     0.51|    8.91|
|Misc              |    36.92|   56.82|
|Misc Lender       |    69.80|  158.43|
|Other             | 23319.00| 5151.45|
|Utility           |    53.14|   55.50|

Essentially, these data show how often (over 13 years from 2001-2014) the owners of individuals' debt in Miami-Dade County, Fl sued those individuals. It's sorted by type of debt.

**Your task**: In 2000, debt buyers (firms that buy debt to collect) sued individuals 59.8 times. This is your `x`. Follow the steps of hypothesis-testing to determine whether this is **significantly different** from the norm over the subsequent 13 years, using a *z*-test and the above information.  This time, use a cut-off of $p=.01$, so there is only a significant difference if $p<.01$. 



