---
title: "Chi-square and factorial ANOVA: Exercises"
author: ""
date: "November 16/19, 2020"
output: 
  html_document:
    self_contained: yes
---

This document is meant to be used to practice after you have completed the tutorial for today's lab. Make sure to put your name as the author of the document, above! 

If you intend to work on these exercises while referring to the tutorial, there are instructions on the [wiki](https://github.com/jdbest/psychRstats/wiki/Run-a-tutorial-and-exercise-simultaneously) on how to do so. You may also want to refer to past labs. Don't forget that previous labs are linked to on the main [labs website](https://jdbest.github.io/psychRstats/labs.html).

## Objectives

In the tutorial, we learned about using `lm()` and `anova()` for factorial ANOVA, about plotting bars against one another using `geom_col()` with `position = position_dodge()` in the parentheses, and about using the `table()` function with `chisq.test()`. In the exercises, you'll use those to practice one of each kind of test.

You can find a completed version of these exercises at <https://jdbest.github.io/psychRstats/answers.html>

Don't forget to (a) save and (b) knit the document frequently, so you'll keep track of your work and also know where you run into errors. 

## Loading packages

As always, you _must_ load packages if you intend to use their functions. Run the following code chunk to load necessary packages for these exercises. 

```{r loadpackages, message=FALSE}
library(tidyverse)
```

## Today's work

### Chi-squared test of independence

Imagine that we get some data on rolling a six-sided die 150 times---just run the chunk below (run all of the lines at once). You'll have 150 numbers between 1 and 6, representing your throwing a die over and over again.

```{r}
# run this whole chunk:
set.seed(42)
dicerolls <- data.frame(
  rolls = sample(c(1:6), 150, replace = TRUE, prob = c(.3,.1,.1,.1,.1,.3))
)
```

1. Run a chi-squared test to test if the die is fair. Take a look at the data, first. Is it tabulated in the right way? What would a p-value less than .05 mean here? 



2. Take a look at the code that "tossed" the die 150 times. Pull the probabilities from there (under the `prob` argument) and see if when you set that as your `p` in the chi-squared test, it changes the results. Why // why not?



3. You can also plot the coin tosses. Use the `ggplot()` function on the data after you've run `group_by()` (grouping by the one column in it) and %>% to `summarize()` to get the `n()` values (i.e., get the frequencies, and then use `ggplot()`.)



### Chi-squared test of goodness of fit

The data below is a "contingency table" (you don't need to run `table()` since it's already one!). It contains 13 house-tasks and their distribution in heterosexual couples: rows are the different tasks; values are the frequencies of the tasks done: by the wife only, by the husband only, by alternating partners, or together. Run the whole code chunk below:

```{r}
contingency.table <- structure(list(Wife = c(156L, 124L, 77L, 82L, 53L, 32L, 33L, 
12L, 10L, 13L, 8L, 0L, 0L), Alternating = c(14L, 20L, 11L, 36L, 
11L, 24L, 23L, 46L, 51L, 13L, 1L, 3L, 1L), Husband = c(2L, 5L, 
7L, 15L, 1L, 4L, 9L, 23L, 75L, 21L, 53L, 160L, 6L), Jointly = c(4L, 
4L, 13L, 7L, 57L, 53L, 55L, 15L, 3L, 66L, 77L, 2L, 153L)), class = "data.frame", row.names = c("Laundry", "Main_meal", "Dinner", "Breakfeast", "Tidying", "Dishes", "Shopping",  "Official", "Driving", "Finances", "Insurance", "Repairs", "Holidays"))
chores <- structure(list(chore = structure(c(1L, 1L, 1L, 1L, 2L, 2L, 2L, 
2L, 3L, 3L, 3L, 3L, 4L, 4L, 4L, 4L, 5L, 5L, 5L, 5L, 6L, 6L, 6L, 
6L, 7L, 7L, 7L, 7L, 8L, 8L, 8L, 8L, 9L, 9L, 9L, 9L, 10L, 10L, 
10L, 10L, 11L, 11L, 11L, 11L, 12L, 12L, 12L, 12L, 13L, 13L, 13L, 
13L), .Label = c("Laundry", "Main_meal", "Dinner", "Breakfeast", 
"Tidying", "Dishes", "Shopping", "Official", "Driving", "Finances", 
"Insurance", "Repairs", "Holidays"), class = "factor"), who = c("Wife", 
"Husband", "Alternating", "Jointly", "Wife", "Husband", "Alternating", 
"Jointly", "Wife", "Husband", "Alternating", "Jointly", "Wife", 
"Husband", "Alternating", "Jointly", "Wife", "Husband", "Alternating", 
"Jointly", "Wife", "Husband", "Alternating", "Jointly", "Wife", 
"Husband", "Alternating", "Jointly", "Wife", "Husband", "Alternating", 
"Jointly", "Wife", "Husband", "Alternating", "Jointly", "Wife", 
"Husband", "Alternating", "Jointly", "Wife", "Husband", "Alternating", 
"Jointly", "Wife", "Husband", "Alternating", "Jointly", "Wife", 
"Husband", "Alternating", "Jointly"), number = c(156L, 2L, 14L, 
4L, 124L, 5L, 20L, 4L, 77L, 7L, 11L, 13L, 82L, 15L, 36L, 7L, 
53L, 1L, 11L, 57L, 32L, 4L, 24L, 53L, 33L, 9L, 23L, 55L, 12L, 
23L, 46L, 15L, 10L, 75L, 51L, 3L, 13L, 21L, 13L, 66L, 8L, 53L, 
1L, 77L, 0L, 160L, 3L, 2L, 0L, 6L, 1L, 153L)), row.names = c(NA, 
-52L), class = c("tbl_df", "tbl", "data.frame"))
```

Let's plot the data from `chores`, first. This is not the contingency table---take a look.

4. Use `ggplot()` to graph `chores`, with `x` being who does which chore, `y` being the number of people who responded which way, and `fill` being the chore itself. 



5. Make some informed guesses: in this data, are chores done evenly or unevenly?

> 

6. Use the `chisq.test()` function to test your hypothesis based on the `contingency.table` ... table.



7. Write up the results of the test, being sure to describe whether it shows independence or not. 

> 

#### Simulating p-values

Generally speaking, chi-squared tests are actually quite bad at getting a *p*-value. One way that R gets around this is by "simulating" *p*-values---running random selections of the data and considering whether *that* would be likely to happen by chance. This is called the Monte Carlo method. 

You can try it with any of your chi-squared tests by adding two arguments: setting `simulate.p.value` equal to TRUE, and choosing how many replications, B, you want, by setting `B` equal to some number (I recommend 10000). 

8. Do that with the `chisq.test()` for the `contingency.table`.



9. How does the *p*-value change?

> 

(For the record, this is why we didn't get into the `pchisq()` function.)

### Factorial ANOVA

Let's just do this based on the `penguins` data since we know it already. Load it by running the following chunk:

```{r}
library(palmerpenguins)
data(penguins)
```

10. Drop the penguins for whom sex is not known, using `filter()` and `! is.na()`. Be sure to save the correctly filtered data back to `penguins`; you'll now have 333 rows instead of 344.



11. Is there an interaction between penguin `sex` and the `island` they live on in predicting `body_mass_g`? Use `lm()` and then `anova()` on the model.



12. Write up the results.

> 

13. Graph the results. Use both a `geom_col()` and a `geom_point() + geom_line()`. For both, be sure to find the **means** before graphing. Do also include a _theme_ and _labels_ for the axes.

For `geom_col()`:
1. Add on error bars using the SEM.
2. Don't forget to include position = position_dodge()
3. In the aes(), make sure to set `fill` equal to one of your independent variables.
4. For the `geom_errorbar()`, add the following code so they line up with the bars (but not inside the `aes()`: `position = position_dodge(.9)`)



For the combination of `geom_line()` and `geom_point()`:
1. In the main `aes()` tag, set `color` equal to one of your independent variables, instead of `fill`. Also set `group` equal to the same variable---this will make the lines connect the correct points.
2. Add on error bars using the SEM---almost identical to the `geom_col()`, but without the `position = position_dodge()` bit.



14. Which of these better describes the data?

> 