---
title: "Lab 04: Hypothesis-Testing in R"
author: "Justin Dainer-Best"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: lumen
    self-contained: true
runtime: shiny_prerendered
tutorial:
  id: "pRs-04"
  version: 0.3
---

```{r setup, include=FALSE}
library(psychRstats)
library(learnr)
library(tidyverse)
library(gradethis)

tutorial_options(exercise.timelimit = 200, exercise.checker = gradethis::grade_learnr)

friends <- structure(list(height = c(60, 60, 74, 62, NA, 65, 65, NA, NA, 
67, NA, 67, 69, 60, 71, 69, NA, 60, 64, 60.5, 59, 61.5, 65, 63, 
NA, 65, 62, 63, 68, 67, 68, 68, 64, 67, 68, 63, 66, 64, 67, 68.5, 
66, 67, 60, 72, 66, 73, 65, 67, 63, 66, 71, 64, 65, 70, 62, 71, 
64, 63, 66, 68, 64, 70, NA, 64, 67, NA, 64, NA, 67, 69, 65, 60, 
61, 65, 69, 6, 65, 73, 65, 73, 67, 5.6, 6, 6.2), tvhours = c(4, 
1, 0, 0, NA, 5, 4, NA, NA, 10, NA, 0, 0, 15, 8, 4, NA, 2, 10, 
4, 0, 5, 4, 3, NA, 7, 5, 14, 0, 6, 3, 0, 8, 0, 1, 10, 14, 15, 
6, 4.5, 17, 7, 10, 7, 1, 3, 2, 25, 14, 5, 5, 2, 3, 2, 2, 7, 7, 
4, 7, 10, 3, 5, NA, 20, 14, NA, 1, NA, 8, 3, 7, 0.5, 8, 3, 1, 
20, 2, 1, 15, 14, 5, 1, 0, 4), gender = c("female", "female", 
"male", "female", NA, "female", "female", "female", NA, NA, "male", 
"female", "non-binary", "female", "male", NA, "male", "female", 
NA, "female", "agender", "male", "male", "female", NA, "female", 
"female", "female", "female", "female", "male", "male", "female", 
"male", "male", "female", "male", "female", "male", "male", "female", 
"female", "female", "male", NA, "male", "female", NA, "female", 
"non-binary", "male", "female", NA, "male", "female", "male", 
"female", "female", "male", "male", "female", "male", "female", 
"female", "female", "female", "female", "female", "female", "female", 
"female", "female", "female", "female", "female", "male", "female", 
"male", "female", "male", "female", "female", "male", "male")), row.names = c(NA, 
-84L), class = c("tbl_df", "tbl", "data.frame")) %>% na.omit()
heights <- friends %>%
  mutate(height = ifelse(height < 10, 
                         floor(height) * 12 + (height - floor(height)),
                         height))
meanheight <- mean(heights$height)
sdheight <- sd(heights$height)
x <- 6*12 + 5
z <- (x - meanheight) / sdheight
```

## Introduction
###
Today's lab is focused on these objectives:

* Learning how to implement a basic hypothesis test in R
* Learning some of the commands that allow you to get probabilities in R
* Practicing the data visualization skills you learned in the previous lab

Let's be clear from the get-go: the methods we're using today aren't very good. We're using data from a small sample and pretending it's from a population. That's fine, because we're learning here---but in the future, you **should not use a *z*-test to run analyses**. 

### Research questions

Today, we'll be using your `friends` data for a few simple questions:

1. Would someone who was 6'5" be considered to have a "significantly different" height among Bard students?
2. Would they be considered "significantly **tall**" compared to other Bard students?
3. Would someone who doesn't watch any TV be considered a statistically-significant outlier among Bard students? 
4. Would that person be considered to watch significantly **less** TV than other Bard students? 

You'll note that (2) and (4) are *one-tailed hypotheses*.

## Step 1
###
***Step 1: Restate question as a research and null hypothesis***

We asked: Would someone who was 6'5" be considered to have a "significantly different" height among Bard students?

```{r quiz1, echo=FALSE}
quiz(
  question("What is the framing of this for the research hypothesis?",
           answer("People who are as tall as this person are different from Bard students as a whole", correct = TRUE), 
           answer("People who are as tall as this person are taller than Bard students as a whole"), 
           answer("People who are as tall as this person are shorter than Bard students as a whole"),
           answer("People who are as tall as this person are no different from Bard students as a whole")
           ),
  question("What is the framing of this for the null hypothesis?", 
           answer("People who are as tall as this person are different from Bard students as a whole"), 
           answer("People who are as tall as this person are taller than Bard students as a whole"), 
           answer("People who are as tall as this person are shorter than Bard students as a whole"),
           answer("People who are as tall as this person are no different from Bard students as a whole", correct = TRUE)
           )
)
```

### Statistical Framing

Ultimately, we're interested in means, as you'll remember---the statistical framing of these hypotheses. How do the means of people like this person compare to the means of students at Bard. We can frame the research hypothesis as 

$\mu_{\mathrm{Bard~students~of~this~height}}\neq\mu_{\mathrm{Bard~students~in~general}}$ 

and the null hypothesis as 

$\mu_{\mathrm{Bard~students~of~this~height}}=\mu_{\mathrm{Bard~students~in~general}}$

## Steps 2 and 3
###
***Step 2: Determine the characteristics of the comparison distribution***

***Step 3: Determine the sample cutoff score to reject the null hypothesis***

In this case, we're continuing to use the distribution of *z*-scores.

```{r quiz2, echo=FALSE}
quiz(
  question(
    "Which of the following is/are true about a z-score distribution? (Select all that apply)",
    answer("We assume that it is normally-distributed", correct = TRUE),
    answer("It has a mean of 0", correct = TRUE),
    answer("It has a standard deviation of 1", correct = TRUE),
    answer("It is symmetrical", correct = TRUE),
    incorrect = "Actually, all of these are true!"
  ),
  question(
    "What is the cut-off score for the z-distribution when using a cutoff of p = .05?",
    answer("The cut-off depends on the data"), 
    answer("The cut-off is +/- 1.64"),
    answer("The cut-off is +/- 1.96", correct = TRUE),
    answer("The cut-off is +/- 2.57")
  )
)
```

With those two questions, we've gotten the hang of Steps 2 and 3 for the first research question!

### R cutoff and `qnorm()`

R can actually determine that cutoff of 1.96 for us with a function called `qnorm()`. It's not something we'll use a lot after today, but it's a function that takes the percentiles from the normal distribution and prints out what *z*-values correspond to them. 

Remember how in class we discussed the fact that what we actually want to get the most extreme 5% of scores is the bottom 2.5% of scores and the top 2.5% of scores? Something like this: 

```{r p1, echo=FALSE}
ggplot(data = data.frame(x = c(-3, 3)), aes(x)) +
  geom_area(stat = "function", fun = dnorm, 
            fill = "red", xlim = c(-3, -1.96)) +
    geom_area(stat = "function", fun = dnorm, 
            fill = "red", xlim = c(1.96, 3)) +
  stat_function(fun = dnorm, size = 1.2, colour = "black") + 
  ylab("") +
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous(breaks = seq(-3, 3, 1), 
                     name = "") + 
  theme_void() + 
  theme(panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(), 
        axis.line = element_line(size = 1),
        axis.title = element_text(size = 24),
        axis.text = element_text(size = 20)) +
  annotate("text", x = -2.3, y = .01, label = "2.5%", color = "white") +
  annotate("text", x = 2.3, y = .01, label = "2.5%", color = "white")
```

Well, we can also think of those as cut-offs for the lowest 2.5% of scores and the highest 2.5% of scores---something like this. (The 97.5% is just $100\% - 2.5\%$.)

```{r p2, echo=FALSE}
ggplot(data = data.frame(x = c(-3, 3)), aes(x)) +
  geom_area(stat = "function", fun = dnorm, 
            fill = "red", xlim = c(-3, -1.96)) +
    geom_area(stat = "function", fun = dnorm, 
            fill = "red", xlim = c(1.96, 3)) +
  stat_function(fun = dnorm, size = 1.2, colour = "black") + 
  ylab("") +
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous(breaks = seq(-3, 3, 1), 
                     name = "") + 
  theme_void() + 
  theme(panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(), 
        axis.line = element_line(size = 1),
        axis.title = element_text(size = 24),
        axis.text = element_text(size = 20)) +
  annotate("text", x = -1.96, y = dnorm(-1.96), label = " 2.5%", size = 5, hjust = 0) +
  annotate("text", x = 1.96, y = dnorm(1.96), label = "97.5%", size = 5, hjust = 1)
```

### Using `qnorm()` to find the cut-off

So to use the `qnorm()` function to get what *z*-value corresponds to each of those, we just need to ask it to give us the *z*-score for 2.5% and 97.5%---using the basic `c()` combine function: `c(.025, .975)`. 

```{r}
qnorm(c(.025, .975))
```

We usually round to two digits in this class, and there's a `round()` function (we tell it how many digits to round **to**). Run the code:

```{r qnorm, exercise=TRUE}
qnorm(c(.025, .975)) %>% round(digits = 2)
```

Yes, that's exactly the $\pm1.96$ we've been using! 

## Step 4

### Step 4: Determine your sample's score

Now, we haven't played with our data yet. The information on heights is saved in the data-frame `friends`. 

Plot a histogram of the data. It might be worth playing around with `ggplot2` (which is loaded), but you can do it with the base plot of `hist()` if you'd like. Your x-variable here will be `height`

```{r h1, exercise = TRUE}

```

```{r h1-solution}
ggplot(friends, aes(x = height)) +
  geom_histogram(binwidth = 2) +
  theme_light()
```

What do you notice about this data? If you said "there are some obvious outliers," you're correct.

### Boxplot

Try using a boxplot plot to see the outliers better. (Remember, that's `geom_boxplot()`.) Make your `x` `height` and your `y` `gender.` No need for a theme or labels---just a quick plot. (*Note: I'm including gender here because you'd expect to see a height difference. We'll talk more about gender at a later date.*)

```{r h2, exercise = TRUE}

```

```{r h2-solution}
ggplot(friends, aes(x = height, y = gender)) +
  geom_boxplot()
```

```{r h2-check, echo=FALSE}
grade_code(correct = "Okay, seems pretty clear to me: some of your friends who responded to the survey seem to have misinterpreted the question. They gave their height in feet instead of inches.")
```

### Problems with heights

Since some folks gave their height in feet instead of inches, if you wanted to exclude them with `filter()`, how would you do that? I'd filter to get a `height` of *more than* 10. (Inches.) Remember that the number should **not** have quotation marks, nor should the variable.

Give it a try below, and then re-run the boxplot. Delete the comments (starting with #) below, and then create a `ggplot()` with a `geom_boxplot()` again. Be careful to use your new data frame and not the old one.

```{r exclude, exercise=TRUE}
tallfriends <- friends # filter here
# ggplot() here +
# geom_boxplot()
```

```{r exclude-solution}
tallfriends <- friends %>% filter(height > 10)
ggplot(tallfriends, aes(x = height, y = gender)) +
  geom_boxplot()
```

```{r exclude-check}
grade_code(correct = "Yes, your boxplot looks more normal now that you've filtered to only height in inches. ")
```

### Fixing that data

Side-step for a moment with me: We can fix those data. First, let's take a look: just run the following code to see what the confusing heights are. 

```{r checkheight, exercise=TRUE}
friends %>% filter(height < 10)
```

Looks to me like those are reasonable heights, if they were instead in feet-inches format... We could fix them! 

###

To fix them, run the following code. Read the comments to make sense of it. We're introducing two new functions: 

1. The `mutate()` function from `dplyr` which, well, "changes" the columns from a data frame. It takes the name of a column (here, `height`) and then changes it according to whatever is on the right of the equal sign. 
2. The `ifelse()` function, which checks a condition and, if it's true, does the first thing; if false, the second. (IF, ELSE.) Argument one: the condition (here, `height < 10`---is the value for each row of height less than 10?). Argument two: what to do if the condition is TRUE (here, a little math assuming that the 5.6 means 5'6"---adding the `5 * 12` to the 6"). Argument three: what to do if the condition is FALSE---if the height isn't less than 10 (here, just keep the same `height` value).

This will have the new data-frame heights, which has things working correctly now. Once you've run this, no values of `heights` will be less than 10. We've "fixed" it.

```{r fixheights, exercise=TRUE}
heights <- friends %>%
  mutate(height = ifelse(height < 10, 
                         floor(height) * 12 + (height - floor(height)),
                         height))

heights %>% filter(height < 10)
```

### Determine the *z*-score

Great! Now comes the "easy" part. We know how to determine a *z*-score already. Let's do it. Remember that $z=\frac{X-M}{SD}$ or (in population terms) $z=\frac{X-\mu}{\sigma}$. So our sample---from which we'll estimate the mean and standard deviation---will be **students in this survey**. You'll see soon enough that we normally make corrections for the fact that this is only an estimate... but for today, let's just calculate the `mean()` and `sd()` of the `height` column from the `heights` data frame, and pretend that those are for a total population. 

In the below field: 

1. Calculate the `mean()` of `heights$height` and assign it (`<-`) to a new variable called `meanheight`
2. Calculate `sd()` of `heights$height` and assign it (`<-`) to a new variable called `sdheight`
3. Assign the height of our 6'5" person (converted to inches) to a new variable called `x`
4. Calculate the *z*-score for them using those new variables, and assign **that** to a new variable called `z`. Don't forget to use parentheses to make clear what happens first.

As you go, I encourage you to make sure the code is working piece-by-piece (by hitting "Run code"), before submitting your answer.

```{r getscore1, exercise=TRUE}

```

```{r getscore1-solution}
meanheight <- mean(heights$height)
sdheight <- sd(heights$height)
x <- 6*12 + 5
z <- (x - meanheight) / sdheight
```

```{r getscore1-check}
grade_code(correct = "Yes, exactly. You'll want to calculate the mean and sd, make sure your x is in the same units (inches) as those, and then calculate z.")
```

## Step 5

### Step 5: Decide whether or not to reject the null hypothesis

You'll probably notice that unless you explicitly asked for the value of *z*, you haven't seen it---step 5 is where we think about it.

Try running the following to check in: what's the value of *z*?

```{r zcheck, exercise=TRUE}
z
```

### Making decisions

As we discussed in Steps 2 and 3, our cut-off is $\pm1.96$. 

```{r quiz3, echo=FALSE}
quiz(
  question(
    "Based on the z-score and the cutoff above, should we reject the null hypothesis?",
    answer("Yes, because the z-score is higher than our cutoff", correct = TRUE),
    answer("Yes, because the z-score is lower than our cutoff"),
    answer("No, because the z-score is neither higher than the cutoff nor lower than the negative cutoff"),
    post_message = paste0("This is important, so discuss with your group if you're still not sure. ",
                          "We can reject the null hypothesis---that there is no difference! ",
                          "This is because the z-score is HIGHER than the cutoff of +1.96.")
  )
)
```

###

We can simply compare to our cutoff---is it more than $+1.96$ or less than $-1.96$?

```{r comparison, exercise=TRUE}
z > 1.96
z < -1.96
```

We can actually do this in one line with the "or" operator, the straight pipe `|`

```{r}
z > 1.96 | z < -1.96
```

But let's go ahead and calculate more than just yes or no, using some useful R functions.

###

Again, like the `qnorm()` function, there's a corresponding one to go the other way---avoiding the *z*-table. It's called `pnorm()` and it prints the percentile for a given *z*-score. Print it out here:

```{r pnorm, exercise=TRUE}
pnorm(z)
```

Turns out that having a height of 6'5" *in our sample* would be extremely unlikely---such a person would fall in the top 99.79% of the sample. 

Compare to a *z*-table (where we would need to add 50% to the "% Mean to z" column or subtract the "% in Tail" column from 100%):

```{r echo=FALSE}
tibble(
  z = seq(2.8, 2.9, by = .01),
  `% Mean to z` = round(100*(pnorm(z, lower.tail = TRUE)-.5), 2),
  `% in Tail` = round(100*pnorm(z, lower.tail = FALSE), 2)
) %>%
  knitr::kable(format = "html")
```

### *p*-value

Let's think about the *p*-value---the actual percentile of this score based on our distribution. How rare is it?

How can we get the actual *p*-value? Turns out that we can make a claim about how rare this height (and this *z*-value) actually is. We'll use the same function `pnorm()`. Because we asked a two-tailed question, we have to ask this in a two-tailed sort of way: what is the probability of getting a value higher than the positive version of this value, or lower than the negative version? We add those together to get the *p*-value: 

```{r}
pnorm(z, lower.tail = FALSE) + pnorm(-z, lower.tail = TRUE)
```

So, we can say that our actual $p=$ `r round(pnorm(z, lower.tail = FALSE) + pnorm(-z, lower.tail = TRUE), 3)`. Which, **yes**, is definitely smaller than $p=.05$, so we can confirm here that $p<.05$.

You won't need to do this calculation with `pnorm()` on your own, but I may walk you through it again...

## A one-tailed test
###
We can do re-run this as a one-tailed test much more quickly now that we have the basics! Let's take the same person---6'5" in height---and return to this question: Would they be considered "significantly **tall**" compared to other Bard students? 

This is a one-sided test---we don't care whether they're on the short side because we know they are not.

### Step 1: Restate question as a research and null hypothesis

```{r quiz4, echo=FALSE}
quiz(
  question("What is the framing of this for the research hypothesis?",
           answer("People who are as tall as this person are different from Bard students as a whole"), 
           answer("People who are as tall as this person are taller than Bard students as a whole", correct = TRUE), 
           answer("People who are as tall as this person are shorter than Bard students as a whole"),
           answer("People who are as tall as this person are no different from Bard students as a whole")
           ),
  question("What is the framing of this for the null hypothesis?", 
           answer("People who are as tall as this person are different from Bard students as a whole"), 
           answer("People who are as tall as this person are the same height or taller than Bard students as a whole"), 
           answer("People who are as tall as this person are the same height or shorter than Bard students as a whole", correct = TRUE),
           answer("People who are as tall as this person are no different from Bard students as a whole"),
           incorrect = "Be careful here! With a one-sided test, or one-tailed test, we are including any result which is not that this person is taller as the null."
           )
)
```

### Statistical framing

We can frame the research hypothesis in statistical terms as 

$\mu_{\mathrm{Bard~students~of~this~height}}>\mu_{\mathrm{Bard~students~in~general}}$ 

and the null hypothesis as 

$\mu_{\mathrm{Bard~students~of~this~height}}\leq\mu_{\mathrm{Bard~students~in~general}}$

Remember, that it's not just LESS THAN---it's LESS THAN **OR EQUAL TO**.

### Steps 2 and 3

***Step 2: Determine the characteristics of the comparison distribution***

***Step 3: Determine the sample cutoff score to reject the null hypothesis***

We're still using the *z*-distribution, but now we're only interested in scores that fall in the upper 5% of the distribution---where 95% of the scores fall below:

```{r p3, echo=FALSE}
ggplot(data = data.frame(x = c(-3, 3)), aes(x)) +
    geom_area(stat = "function", fun = dnorm, 
            fill = "red", xlim = c(qnorm(.95), 3)) +
  stat_function(fun = dnorm, size = 1.2, colour = "black") + 
  ylab("") +
  scale_y_continuous(breaks = NULL) +
  scale_x_continuous(breaks = seq(-3, 3, 1), 
                     name = "") + 
  theme_void() + 
  theme(panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(), 
        axis.line = element_line(size = 1),
        axis.title = element_text(size = 24),
        axis.text = element_text(size = 20)) +
  annotate("text", x = qnorm(.95), y = dnorm(qnorm(.95))/2, label = "95%", size = 5, hjust = 1)
```

Again, we can use the `qnorm()` function to ask for the upper 95% cut-off:

```{r}
qnorm(.95)
```

Yep, that's what we discussed in class! Our cut-off is $+1.64$. 

### Step 4: Determine your sample's score

You already determined the score!---here, it's:

```{r}
x
```

And the corresponding *z*-score, which we can use to compare, is:

```{r}
z
```

### Step 5: Decide whether or not to reject the null hypothesis

To decide whether or not to reject the null hypothesis, we don't actually need to think too hard---if something was statistically significant for a two-tailed test, a one-tailed test will always be statistically significant at the same *p*-value. (One-tailed tests, as we discussed in class, can be "more lenient", because they're only "looking" at one tail of the distribution.)

We can simply compare to our cutoff. Hit run code:

```{r comparison2, exercise=TRUE}
z > 1.64
```

Okay, ready to get more than just the "yes or no" answer?

### 

Rather than needing to calculate the *p*-value at the top and bottom, we can just do it for the top, this time---since we only expected it to be on the top!

```{r pnorm2}
pnorm(z, lower.tail = FALSE)
```

So we can say that $p=$ `r round(pnorm(z, lower.tail = FALSE), 3)`, and again, confirm that in this case $p<.05$. 

Any questions? Now's a great time to check in and make sure you understand everything. 

You'll answer questions 3 and 4 from above---about number of hours of TV---in the exercises. 

## Wrap up

Once you've finished, go back to the website for today's exercises, open the exercise directly, or run the following code (from your primary working directory):

```{r, eval=FALSE}
psychRstats::run_exercise("04-exercise")
```